---
output: github_document
---

<!-- README.md is generated from README.rmd. Please edit that file -->

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
library(plt)
```

# plt

Persistence landscapes are a vectorization of persistence data/diagrams that have useful statistical properties including linearity and an inner product.[^pl]
This is an R package interface to a C++ library to efficienctly compute and calculate with persistence landscapes.[^plt]

[^pl]: Bubenik P (2015) "Statistical Topological Data Analysis using Persistence Landscapes". _JMLR_ **16**(3):77--102. <https://jmlr.csail.mit.edu/papers/v16/bubenik15a.html>
[^plt]: Bubenik P & Dłotko P (2017) "A persistence landscapes toolbox for topological statistics". _Journal of Symbolic Computation_ **78**(1):91--114. <https://www.sciencedirect.com/science/article/pii/S0747717116300104>

## Installation

Until the package is on CRAN, use **remotes** to install the package from the GitHub repository as follows:

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("corybrunson/plt", build_vignettes = TRUE)
```

Alternatively---and especially if you want to contribute---you can clone or download the code repository and, from within the directory, install the package from source:

```{r, eval=FALSE}
devtools::install()
```

You should now be able to load the package normally from an R session:

```{r, eval=FALSE}
library(plt)
```

## Quickstart Guide

The **plt** package supports various operations involving persistence landscapes:

* Compute persistence landscapes from persistence data
* Perform Hilbert space operations (scaling, addition, inner product) plus some additional queries and transformations (extremal values, absolute value, integration) on persistence landscapes
* Plot persistence landscapes

Examples and tests in **plt** rely on other packages to simulate data and to compute persistence diagrams from data:

* **tdaunif** provides functions to sample uniformly from various immersed manifolds.
* **ripserr** and **TDA** provide functions to compute persistence data from point clouds and distance matrices.

**plt** introduces the 'Rcpp_PersistenceLandscape' S4 class, which is exposed using **Rcpp** from the underlying 'PersistenceLandscape' C++ class.
Instances of this class can be created using `new()` but the recommended way is to use `landscape()`.
This function accepts either a single matrix of persistence data or a specially formatted list with the class `'persistence_diagram"`.
The `$pairs` entry of the list is itself a list, of a 2-column matrix of persistence pairs for each homological degree from 0 (`$pairs[[1]]`) to the maximum degree calculated.
The generic converter `as_persistence()` includes methods for outputs from `ripserr::vietoris_rips()` and from `TDA::*Diag()`; it operates under the hood of `landscape(), but we invoke it explicitly here for illustration.

### Calculation

To begin an illustration, we noisily sample 60 points from a figure eight and compute the persistence diagram of the point cloud:

```{r}
set.seed(513611L)
pc <- tdaunif::sample_lemniscate_gerono(60, sd = .1)
pd <- ripserr::vietoris_rips(pc, dim = 1, threshold = 2, p = 2)
print(pd)
```

We the convert the persistence data to the preferred persistence diagram format and inspect some of its features:

```{r}
pd <- as_persistence(pd)
print(pd)
print(head(pd$pairs[[1]]))
print(head(pd$pairs[[2]]))
```

This allows us to compute a persistence landscape---in this case, for the 1-dimensional features.
Here we compute the landscape exactly, which can be cost-prohibitive for larger persistence data:

```{r}
pl1 <- landscape(pd, degree = 1, exact = TRUE)
print(pl1)
```

### Class

The object `pl1` is not an array, but rather an object that encapsulates both the data that encode a landscape and several basic operations that can be performed on it.
This allows us to work with persistence landscapes without worrying about pre-processing their representations.
At any point, the underlying encoding of the landscape can be extracted using `$getInternal()`, which in the case of an exactly calculated landscape returns a list of 2-column matrices, each matrix containing coordinates that define one level of the landscape as a piecewise linear function:

```{r}
print(length(pl1$getInternal()))
print(pl1$getInternal())
```

An alternative, approximate construction computes the value of each level of the landscape at each point on a 1-dimensional grid, ranging from `xmin` to `xmax` at increments of `by`.
A landscape constructed using a discrete approximation is stored as a 3-dimensional array of dimensions (levels, values, 2), with one level per feature (some of which may be trivial) and one value per grid point, stored as $x,y$ pairs along the third dimension.

```{r}
b_ran <- pl_support(pl1)
pl1d <- landscape(pd, degree = 1,
                  xmin = b_ran[[1L]], xmax = b_ran[[2L]], by = 0.05)
print(dim(pl1d$getInternal()))
print(pl1d$getInternal())
```

Exactly computed landscapes can be converted to discrete landscape objects, but the other direction is not well-defined.
Below, we view a portion of the discretized exact landscape, using the default bounds and resolution given to `pl1`:

```{r}
# default conversion to discrete uses `by = 0.001`
pl1_ <- pl1$discretize()
print(dim(pl1_$getInternal()))
# print first 12 x-coordinates
pl1_$getInternal()[, seq(230L, 270L), , drop = FALSE]
```

We can also specify the bounds and the resolution of the discretization:

```{r}
pl1 <- pl_delimit(pl1, xmin = 0, xmax = 1, by = 0.1)
pl1_ <- pl_discretize(pl1)
pl1_$getInternal()
```

### Visualization

**plt** provides a `plot()` method for the 'Rcpp_PersistenceLandscape' class.
It uses **grDevices** to build color palettes, and as such its default palette is viridis; but the user may supply the name of a recognized palette or a sequence of colors between which to interpolate:

```{r}
n_levs <- max(pl_num_levels(pl1), pl_num_levels(pl1d))
par(mfrow = c(2L, 1L), mar = c(2, 2, 0, 2))
plot(pl1, palette = "terrain", n_levels = n_levs, asp = 1)
plot(pl1d, palette = "terrain", n_levels = n_levs, asp = 1)
par(mfrow = c(1L, 1L), mar = c(5.1, 4.1, 4.1, 2.1))
```

### Vector Operations



### Inner Product



## Acknowledgments

### Precursors

The C++ library is adapted from [Paweł Dłotko's Persistence Landscape Toolbox](https://www2.math.upenn.edu/~dlotko/persistenceLandscape.html).
It was originally adapted and ported to R in [Jose Bouza's **tda-tools** package](https://github.com/jjbouza/tda-tools).

### Resources

Development of this package benefitted from the use of equipment and the support of colleagues at the [University of Florida](https://www.ufl.edu/), especially [Peter Bubenik's research group](https://people.clas.ufl.edu/peterbubenik/researchgroup/).

### Contribute

Bug reports, unit tests, documentation, use cases, feature suggestions, and other contributions are welcome. See the [CONTRIBUTING](https://github.com/corybrunson/plt/blob/main/CONTRIBUTING.md) file for guidance, and please respect the [Code of Conduct](https://github.com/corybrunson/plt/blob/main/CODE_OF_CONDUCT.md).

### Cite

If you use **plt** in published work, please include a citation following `citation("plt")`.
