---
title: "Exploratory and Predictive Analysis"
author: "James Golabek, Aidan Bryant, Jason Cory Brunson, Johnathan Bush, Iryna Hartsock, Peter Bubenik"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Exploratory and Predictive Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette, we will show how {plt} supports the computation of persistence landscapes[^1], and both exploratory and predictive analysis. The construction of persistence landscapes has useful properties for machine learning and hypothesis testing. 

# Basic use of TDA in Machine Learning

**Topological data analysis** (TDA)[^2] involves the use of algebraic topology to analyze data. The goal is to measure global topological features. A typical procedure is to: 

1. Represent a sample of data as a point cloud.
2. Construct a filtered simplicial complex on the point cloud.
3. Obtain the persistence module by taking the simplicial homology of the filtration.
4. Visualize/vectorize the persistence module.
5. Apply hypothesis tests and machine learning tools.

This vignette will include dimension reduction and supervised classification as part of the applications of TDA.

# Interoperability

{plt} is designed to be used with other TDA packages such as those loaded below. 

The [{tdaunif}](https://github.com/tdaverse/tdaunif) package provides tools to take uniform random samples from simple manifolds and is commonly used for testing TDA tools.
The [{TDA}](https://cran.r-project.org/package=TDA) package contains various tools for TDA, including functions to compute persistence data from point clouds and distance matrices. We use the `alphaComplexDiag()` function to obtain the birth and death values of the alpha complex persistence module.

```{r,warning=FALSE,message=FALSE}
# for manifold samples
library(tdaunif)
# for alpha complexes
library(alphahull)
# for persistent homology
library(TDA)
# for persistence landscapes!
library(plt)
# for support vector machines
library(e1071)
```

# Gather a point set

To begin, we create a simple point cloud on which we will illustrate how to use {plt}. We noisily sample $60$ points in $\mathbb{R}^2$ from a figure eight, the lemniscate of Gerono.

```{r,fig.align = 'center'}
set.seed(120246L)
pc <- sample_lemniscate_gerono(n = 60, sd = 0.05)
plot(pc, asp = 1, pch = 16L, main = "Figure Eight Point Cloud")
```

# Adapting Point Data to Persistence Data

From our point set, we would like to generate more topologically informative data. We construct a **simplicial complex** $K$, which is a set of **simplices**. A *$0$-simplex* is a single point, a *$1$-simplex* is an edge or line segment, a *$2$-simplex* is a triangle, and a *$3$-simplex* is a tetrahedron. 

There are different formulations for constructing simplicial complexes that have their own advantages and disadvantages. We will utilize the **alpha complex** as it captures the underlying structure of the data accurately and is more computationally efficient than other methods. 

By applying the alpha complex with various radii (scale), we create a **filtered simplicial complex** or **filtration**, which is a nested sequence of simplicial complexes:
$$
K_0 \subset K_1 \subset K_2 \subset \cdots\subset K_n,
$$
where $n \in \mathbb{N}$ is the number of simplicial complexes in the filtration. For any pair of adjacent terms (simplicial complexes) in the filtration, the preceding term is a **subcomplex** of the succeeding term. 

# Create a Simplicial Complex

The plots below show the alpha complex at six levels of its distance threshold $\epsilon_\alpha$.  As $\epsilon_\alpha$ increases, the number of connected components and loops changes. However, some topological features, such as the two lobes of the figure eight, persist throughout many values of $\epsilon_\alpha$.

```{r,echo = FALSE,fig.align = 'center'}
par(mfrow = c(3, 2), mar = c(2,2,1,1))

alpha_values <- c(0,0.05,0.1,0.175,0.35,0.4)

for (alpha in alpha_values) {
   plot(ashape(pc,alpha=alpha),col=c(4,1),xlab="",ylab="")
}
```

# Compute Persistence Data

The filtration provides a mathematical structure for analyzing the global topological features of a point cloud. For each simplicial complex within the filtration, we apply **simplicial homology** to classify them using Betti numbers. The **$k^{\text{th}}$ Betti number $\beta_k$** counts the number of $k$-dimensional features of a simplicial complex. **$\beta_0$** counts the number of $0$-dimensional connected components, **$\beta_1$** counts the number of $1$-dimensional loops, and **$\beta_2$** counts the number of $3$-dimensional cavities. 

Applying simplicial homology to the filtration, we obtain a sequence of vector spaces and linear maps called a **persistence module**:
$$
H_k(K_0) \rightarrow H_k(K_1) \rightarrow H_k(K_2) \rightarrow \cdots  \rightarrow H_k(K_n).
$$
The persistence module tracks the *lifetime* of features throughout the *time span* of the filtration. The *birth* value $b$ of a feature is when it first appears in the persistence module. The *death* value $d$ of a feature is when it disappears. A topological feature with a long lifetime $d-b$ is called a *persistent* feature.

```{r}
pd <- alphaComplexDiag(pc, maxdimension = 1)$diagram

# birth and death values have been squared, so take the square root
pd[,c(2,3)] <- sqrt(pd[,c(2,3)])
```

The `pd` object stores the birth and death values for the $0$ and $1$ dimension features across the alpha filtration.

```{r}
head(pd)
```


<!--
### Persistence Barcode
A persistence barcode is a topological summary that visualizes the lifetimes of topological features. The x-axis represents the time span of the filtration. The below barcode shows that two features in homology dimension-one persist longer than the other features in dimension one. These two persistent features correspond to the two holes in $X$.

```{r}
plot(pd, diagLim = c(0, 0.5), barcode = TRUE)
legend(0.3, 25, c('Homology in degree 0','Homology in degree 1'),
col = c(1,2), pch = c(19,2), cex = .8, pt.lwd = 2)
```
-->

# Persistence Diagram

The persistence diagram is a topological summary that visualizes the birth and death times of topological features. The diagram has axes birth & death, and a line $d = b$. For any point $(b,d)$, we have $d-b \geq 0$. Graphically, each point is on or above the $d = b$ line. Points close to or on this line are *non-persistent* features because they die shortly or immediately after being born. 

In the persistence diagram below, three features stand out. Two are $1$-dimensional features distinguished by their distance from the line $d = b$. These persistence features represent the two lobes of the figure eight. We also observe a $0$-dimensional feature with infinite persistence because there must always be the one connected component that includes the entire dataset. As smaller components merge, this final component persists across all scales, since nothing remains to connect with it.

```{r,fig.align = 'center'}
plot(pd, asp = 1, diagLim = c(0,.75))
legend(0.6, 0.4, c('Homology in degree 0','Homology in degree 1'),
col = c(1,2), pch = c(19,2), cex = .8, pt.lwd = 2, bty = "n")
```

#  Landscape Transformation

Now that we have persistence data, we can construct persistence landscapes, which are useful for analyzing topological features in a form suited for statistical methods. Persistence landscapes are built by transforming persistence diagrams into a sequence of "tent" functions that capture the lifetime of topological features.

## Tent Functions
Each point in the persistence diagram corresponds to a topological feature with a birth and death time. We can represent these features using tent functions:

- The base of the tent spans from the feature's birth to its death.
- The peak of the tent is located at the midpoint between birth and death.
- The height of the tent at any point reflects how long the feature persists around that time.

These tent functions are given by:
$$
f_{(b_i, d_i)}(x) = 
\begin{cases} 
x - b_i & \text{if } b_i < x \leq \frac{b_i + d_i}{2} \\ 
-x + d_i & \text{if } \frac{b_i + d_i}{2} < x < d_i \\ 
0 & \text{otherwise} 
\end{cases}
$$

## Levels
To create the persistence landscape, we superimpose these tent functions. At each point on the timeline, the height of the landscape is given by the level of these tents—the largest height at that point. The level provides a layered structure:

- $\lambda_1(t)$ is the level that takes the value of the tallest tent at $t$
- $\lambda_2(t)$ is the level that takes the value of the second tallest tent at $t$, and so on for the remaining less prominent features.
- Generically, $\lambda_k(t)=\text{k}^{\text{th}}$ largest value of $\{f_{(b_i, d_i)}(t) | (b_i, d_i) \in PD\}$ at each $t$ where $PD$ is the persistence diagram.
This superimposition of tents, creating a level of functions, allows us to visualize and analyze the relative significance of topological features. The collection of our levels makes our landscape.



These levels help develop a geometric understanding of persistence landscapes and should be helpful for when we think about the analysis being done. 


To begin working with the persistence landscape framework, {plt} methods demands that our data is in a particular format, `as_persistence()` ensures this, if your data is in one of the recognized formats/classes (see [help file](https://corybrunson.github.io/plt/reference/as_persistence.html)).

This function in {plt} has multiple methods to convert persistence data to a format required for other {plt} functions. It can be applied to a $3$-column matrix (or an object coercible to one) with columns dimension/degree, start/birth, and end/death, though the columns need not be named, or it can convert outputs from persistence data producing functions like `alphaComplexDiag()`.

Note that `as_persistence()` is called in `pl_new()`.

```{r}
pd <- as_persistence(pd)
print(pd)
```

The `pl_new()` function takes in a single matrix of persistence data or a formatted list with the class `persistence_diagram`, and computes the persistence landscape. Below, we compute the landscape for the $1$-dimensional topological features.

Often users will find it useful to only look at one dimension of the data, whether for simplification or if it's a particular dimension of interest.
Below, we filter for only degree-one topological features and the entire domain of the landscape by setting `xmin` and `xmax` to the infimum and supremum, respectively. You can select what window of filtration you want the landscape by changing these parameters. The window must contain the *support* of the landscape, which is the set of points for which the persistence landscape is non-zero.

```{r}
pl1d <- pl_new(pd, degree = 1, xby = 0.025)
summary(pl1d)
```

The above function for the "tents" performs a transformation that puts each feature in terms of $\frac{b+d}{2}$ and $\frac{d-b}{2}$, making the graphical representation easier to visually understand. 
Note that this is a discrete persistence landscape so it is generated by sampling a number of points and interpolating to create a continuous function.

```{r,fig.align = 'center'}
n_levs <- max(pl_num_levels(pl1d))
plot(pl1d, main = "Persistence Landscape", n_levels = n_levs, asp = 1)
```


# Mean Persistence Landscape

A convenient way of thinking about the mean persistence landscape is by considering its levels. To define a mean on landscapes, we take the pointwise mean of each level. Hence, the mean $i^{\mathbb{th}}$ level is given by
$$\bar \lambda_i(x) = \frac{1}{N}\sum_{j=1}^{N}\lambda_i ^j(x),$$
where $j$ is the index of the landscape, and $N$ is the number of landscapes. Overlaying these mean levels grants us the mean persistence landscape.

Below, we noisily sample $6$ times from a lemniscate, and then compute the persistence landscape for each sampled point cloud.

```{r,fig.align = 'center'}
set.seed(101137L)
par(mfrow = c(3, 2), mar = c(2,2,1,1))
pl_list_1 <- c()
for (i in seq(6)) {
  pc <- sample_lemniscate_gerono(n = 60, sd = 0.05)
  pd <- alphaComplexDiag(pc, maxdimension = 1)$diagram
  pd[,c(2,3)] <- sqrt(pd[,c(2,3)])
  pl <- pl_new(pd, degree = 1, xby = .01)
  plot(pl)
  pl_list_1 <- c(pl_list_1, pl)
}
```

We then compute the mean persistence landscape of these $6$ persistence landscapes.

```{r,fig.align = 'center'}
plot(pl_mean(pl_list_1), main = "Mean Persistence Landscape")
```

# Hypothesis Testing

In statistics, hypothesis tests are used to weigh the evidence (collected data) to determine whether there is a real effect or phenomenon present in the data, or if its simply a consequence of random sampling or variation. When conducting a hypothesis test, we define a _null hypothesis_ $H_0$ that predicates that no effect is present, and we define an alternative hypothesis $H_A$ that specifies the direction of a present effect. A hypothesis test requires a calculated *test statistic* from the observed data, and an understanding of the test statistic's theoretical probability distribution. A *$p$-value* is the probability of obtaining a test statistic as extreme or more extreme than the observed test statistic, assuming the null hypothesis to be true. And $\alpha$, the *significance level*, is the probability of rejecting the null hypothesis when it is actually true (*type I error*). This value is typically set by the investigator ahead of time, conventionally to 0.05. If the $p < \alpha$, then we _reject_ our null hypothesis, otherwise we _accept_ or _fail to reject_ it.

In TDA, we can use persistence landscapes to derive test statistics in order to conduct hypotheses tests on samples of point cloud data. Our {plt} hypothesis testing functions --- `pl_z_test()`, `pd_z_test()`, and `pl_perm_test()` --- provide the test result as a list containing useful information like the estimated distance between the mean landscapes, the test statistic, the $p$-value, and more. The list is assigned the `htest` class, which comes with an elegant `print()` method.

For our next example, we will show how to test whether two collections of point clouds were sampled from different distributions. First, we create a counterfeit lemniscate sampler, which might be similar enough to that of Gerono to, say, deceive a quality control step. This counterfeit lemniscate is the union of differently-sized circles, while the original has same-sized lobes, and it is sampled with noise from the same distribution.

```{r,fig.align = 'center'}
sample_nonsym_lemniscate <- function() {
  circle1 <- circleUnif(n = 25, r = .50)
  circle2 <- circleUnif(n = 35, r = .75)
  
  circle1 <- cbind(circle1[, 1] + 1, circle1[, 2] + 1)
  circle2 <- cbind(circle2[, 1] + 2.25, circle2[, 2] + 1)
  
  circle1 <- add_noise(circle1, sd = 0.05)
  circle2 <- add_noise(circle2, sd = 0.05)
  Y <- rbind(circle1, circle2)
}
plot(sample_nonsym_lemniscate(), asp = 1, pch = 16, xlab = "x", ylab = "y", main = "Non-symmetric Figure Eight Point Cloud")
```

We noisily sample $6$ times from the lemniscate sampler above, compute the persistence landscape for each point cloud, and then plot the persistence landscapes.

```{r,fig.align = 'center'}
set.seed(194114L)
par(mfrow = c(3, 2), mar = c(2,2,1,1))
pl_list_2 <- c()
for (i in seq(6)) {
  pc <- sample_nonsym_lemniscate()
  pd <- alphaComplexDiag(pc, maxdimension = 1)$diagram
  pd[,c(2,3)] <- sqrt(pd[,c(2,3)])
  pl <- pl_new(pd, degree = 1, xby = .01)
  plot(pl)
  pl_list_2 <- c(pl_list_2, pl)
}
```

We will refer to the sampling distributions, `pl_list_1` and `pl_list_2`, as $P_1$ and $P_2$. Below, we replot the mean persistence landscape of $P_1$ (figure eight) and plot the mean persistence landscape of $P_2$ (non-symmetric figure eight). We denote these mean landscapes as $\bar{P_1}$ and $\bar{P_2}$.
```{r,fig.align = 'center'}
par(mfrow = c(2,1), mar = c(2,2,0.25,2))
plot(pl_mean(pl_list_1))
title("Mean PL of P1", adj = 0.1, line = -1)
plot(pl_mean(pl_list_2))
title("Mean PL of P2", adj = 0.1, line = -1)
```

There is a visual difference between the landscapes, but we haven't quantified the strength of evidence for this difference.
We will use a two sample Z-test and a permutation test to determine if there is a statistically significant difference between $\bar{P_1}$ and $\bar{P_2}$.

## Z-Test

The purpose of a **two sample Z-test** is to test for the difference of means between two samples from two population distributions. The Z-test is a parametric test that requires the two samples to be normally distributed and performs well when you have large sample sizes and known population variances. If these assumptions are not met, then the test is inappropriate.

To conduct this test on persistent landscapes, the test statistic requires a functional, the integral of the mean persistent landscape, to quantify the difference in mean persistence landscapes.  The interval of this integration is determined by the `support` provided from the user. Specifically, users can enter intervals for the `support` parameter to focus on regions where important topological features lie, which would increase computational efficiency. The test statistic is 
$$z = \frac{\bar{P_1}-\bar{P_2}}{\sqrt{\frac{S^2_{P_1}}{n_1}+\frac{S^2_{P_2}}{n_2}}},$$

where the denominator is the standard error for the difference and $z \sim N(0,1)$. For more information, please reference section 3 of Bubenik (2015).   

In our example, the null hypothesis is that there is not a difference in the two mean persistence landscapes, which would imply that they both came from the same population distribution. And the alternative hypothesis is that there is a difference between the mean persistence landscapes. Succinctly, we test $H_0: \bar{P_1}-\bar{P_2} = 0$ $\text{vs}$ $H_A: \bar{P_1}-\bar{P_2} \ne 0$.

We can adjust $\alpha$ by changing the `conf.level`. Here, we will use the default $0.95$ confidence level, so $\alpha=0.05$. Also, the user can adjust the direction of the alternative hypothesis by setting the `alternative` argument to "two.sided" (default), "greater" or "less".

```{r}
(ztest <- pl_z_test(pl_list_1, pl_list_2, alternative = "two.sided", conf.level = 0.95))
```

```{r,echo = FALSE, include=FALSE}
pval1 <- round(ztest$p.value, 3)
```

Since the $p$-value is $`r pval1`$, we reject the null hypothesis. This result suggests that, according to the Z-test, the samples appear to come from the different distributions, though the evidence was not as strong as we might want. This highlights the Z-test's conservative nature when testing for differences between two samples of persistence landscapes, especially when we have small sample sizes.

## Permutation Test

We now perform a **permutation test**, which is a non-parametric hypothesis test. This test is useful for smaller sample sizes and does not require an assumption of distribution of the data. Specifically, the principle assumption of the permutation test is that the observations (in our case, the persistence landscapes) do not come from two distinct population distributions $F_{P_1}$ & $F_{P_2}$, and that they are *completely exchangeable* across the two groups.

The first step of the permutation test is to calculate an initial test statistic. For our data, it will be 
$K = \bar{P_1}-\bar{P_2}$. We build the sampling distribution by rearranging our $12$ samples ($6$ each from $P_1$ and $P_2$) and computing the test statistic for each permutation. We then compare the initial test statistic to the distribution of test statistics we obtained by randomly permuting the data. This helps to determine whether the observed effect is statistically significant. 

The null hypothesis is that $P_1$ and $P_2$ come from the same distribution and the alternative hypothesis being that they do not come from the same population distribution. Succinctly, we test $H_0: F_{P_1} = F_{P_2}$ $\text{vs}$ $H_A:  F_{P_1} \ne F_{P_2}$.

```{r}
(perm <- pl_perm_test(pl_list_1, pl_list_2, complete = TRUE))
```

```{r,echo = FALSE, include=FALSE}
pval2 <- round(perm$p.value, 4)
```

Since the $p$-value is $`r pval2`$, we again reject the null hypothesis. This $p$-value was an order of magnitude smaller than that obtained from the two sample Z-test. These two hypothesis tests are looking at the data in different ways, so one test may detect the topological (or geometric) difference between the samples more effectively than the other.

# Exploratory Dimension Reduction

In this section, we conduct exploratory analysis using variance decomposition.  This is done to reduce the dimensions of our data while retaining as much variance as possible. We will use **principal components analysis** (PCA), which transforms the given variables into a new set of uncorrelated variables called *principal components*. These components are ordered by the amount of variance they capture from the data, with the first component capturing the most variance, and each subsequent component capturing progressively less. We apply PCA to the vectorized forms of the persistence landscapes.

We continue using the lists from our example. We utilize the `pl_to_matrix()` function, which converts our persistence landscapes into matrix format, allowing us to perform general-purpose data analytic operations on them. `prcomp()` performs PCA on our vectorized persistence landscapes.

```{r,fig.align = 'center'}
pl_list <- c(pl_list_1,pl_list_2)
pl_vectors <- pl_to_matrix(pl_list)
pca <- prcomp(pl_vectors)
plot(pca, type = "l", main = "Scree Plot")
summary(pca)
```

In the summary above, we see that about $77.5\%$ of the variance in the original data is captured in the first principal component, and about $13.9\%$ of the variance is captured in the second principal component. So the first and second principal component are sufficient to explain the variance of the data.

To evaluate whether the dimension reduction was successful, we assign labels to our data based on the distribution they were generated from and plot each sample with their corresponding first principal component and second principal component values to see if any clustering or patterns occurred.

```{r,fig.align = 'center'}
data.labels <- c(rep(1,length(pl_list_1)), rep(2,length(pl_list_2)))
plot(pca$x[,1:2], col = data.labels, pch = 17 + (2 * data.labels), asp = 1)
```

As we can see, clustering of samples from the same sample did occur. We observe that the between-group variance is mainly contained in the first principal component. Also, the within-group variance of $P_1$ is smaller than the between-group variance of $P_2$. 

# Classification

We now build a machine learning model to classify the point clouds using their vectorized persistence landscapes. A **support vector machine** (SVM) is a supervised algorithm commonly used for classification tasks. It works by finding the optimal hyperplane that best separates the classes ($P_1$ or $P_2$) in the feature space (where each dimension of the feature space corresponds to one of the elements in these vectors), maximizing the margin between the closest data points of each class, known as support vectors.

Below we train our model using `svm()` and perform $4$-fold *cross validation*. That is, the algorithm splits the data into $4$ groups and as it iterates to using each group as the testing group, it uses the other $3$ groups as the training data. We set `scale = FALSE` so that more persistent features have more weight.

```{r}
svm_model <- svm(pl_vectors,data.labels,scale = FALSE,
                 type = "C-classification", kernel = "linear", cost = 10, cross = 4)
summary(svm_model)
```

The model successfully distinguishes between the different topological structures present in the two persistence landscapes.

# Appendix

A **$k$-simplex** is a $k$-dimensional polytope which is the convex hull of $k+1$ points in Euclidean space in general position (the points $x_1,...,x_{k+1}$ are *affinely independent*, which means that the vectors $x_2-x_1,...,x_{k+1}-x_1$ are linearly independent). The convex hull of a non-empty subset of these points is called the *face* of the simplex. 

A **simplicial complex** is a set $K$ of simplices such that

- the face of any simplex in $K$ is also in $K$; and

- any non-empty intersection of two simplices is a face of both of them.

An **abstract simplicial complex** is a set $K$ of non-empty subsets of a fixed set such that if $A \in K$, then all nonempty subsets of $A$ are also in $K$.

For a finite set of points $X \subset \mathbb{R}^d$, the **Alpha complex** Alpha$(X,s)$ is a simplicial subcomplex of the Delaunay complex of $X$ consisting of simplices of circumradius less than or equal to $\sqrt{s}$. The Alpha complex is defined as:
$$ 
\text{Alpha}(X, r) = \{ \sigma \subset X : \bigcap_{u \in \sigma} R_u(\sqrt{s}) \neq \emptyset \}.
$$

A **filtered simplicial complex** consists of a nested sequence of simplicial complexes:
$$
K_0 \subset K_1 \subset K_2 \subset \;...\subset K_n
$$

Applying homology, we obtain a sequence of vector spaces and linear maps called a **persistence module**:
$$
H_k(K_0) \rightarrow H_k(K_1) \rightarrow H_k(K_2) \rightarrow \;...\rightarrow H_k(K_n)
$$
The images of these maps and compositions of these maps are called **persistence homology** vector spaces.

A **persistence module** $M$ consists of a vector space $M_a$ for all $a \in \mathbb{R}$ and linear maps $M(a\leq b):M_a \to M_b$ for all $a \leq b$ such that $M(a \leq a)$ is the identity map and for all $a \leq b \leq c, \; M(b \leq c) \circ M(a \leq b) = M (a \leq c)$.

Let $M$ be a persistence module and $b_i \leq d_i$ where $b_i$ is some birth value and $d_i$ is some death value. The **betti number** of $M$ is 
$$
\beta^{b_i, d_i} = \dim(\text{im}(M(b_i \leq d_i)))
$$

The **$k^{th}$ Betti number**, $\beta_k$, counts the number of $k$-dimensional features of a simplicial complex. For example, $\beta_0$ counts the number of connected components and  $\beta_1$ counts the number of holes. 

The **rank function**, $\lambda: \mathbb{R}^2 \rightarrow \mathbb{R}$ defined as:
$$
\mathbb{\lambda}(b, d) = 
\begin{cases}
    \mathbb{\beta}^{(b,d)} & \text{if } d \geq b \\
    0 & \text{otherwise}
\end{cases}
$$

Now, we change the coordinate system from (birth, death) scale to the (midlife, halflife) scale for ease of interpretability and analysis. 
We have: 
$$
m = \frac{(b + d)}{ 2} \;  \text{and} \;h = \frac{(d-b)}{2}.
$$
Hence our **rescaled rank function**, $\mathbb{\lambda}: \mathbb{R}^2 \rightarrow \mathbb{R}$, defined as:  
$$
\mathbb{\lambda}(m, h) = 
\begin{cases}
    \mathbb{\beta}^{(m-h,m+h)} & \text{if } h \geq 0 \\
    0 & \text{otherwise}
\end{cases}
$$.

Finally, the **persistence landscape** is a function, $\lambda: \mathbb{N} \times \mathbb{R} \rightarrow \mathbb{R}$. Or it could be thought of as a sequence of functions $\lambda_k: \mathbb{R} \to \mathbb{R}$, where $\lambda_k(t) = \lambda(k,t)$. $\lambda_k(t)$ is defined as: 
$$
\lambda_k(t) = \sup(m \geq 0 | \beta^{t-m, t+m} \geq k)
$$
Essentially, $\lambda_k(t)$ finds the largest interval around t where there at least $k$ persistent features.
 
# References

[^1]: Bubenik, Peter. "Statistical Topological Data Analysis Using Persistence Landscapes." *Journal of Machine Learning Research*, vol. 16, 2015, pp. 77-102.

[^2]: https://people.clas.ufl.edu/peterbubenik/intro-to-tda/ 
